---
title: "Week 3: Importing and Manipulating Data II"
output: html_document
---

Let's read in our data (same as last week)

```{r}
setwd("U:/LOCKED/arthropod_data")#Yours may not be the "U:" drive, make sure you enter the correct letter
commDat=as.data.frame(read.csv("commComp_2000.csv",na.strings=c("","NA")))#The na.strings command means that any blank or NA cells will be given a value of NA. We're making it a data.frame so that it differentiates between strings (i.e. words) and numbers
```

We'll start by learning a couple of techiniques for subsetting the matrix by rows and columns
For example, we can pull out just the records for species in the Hymenoptera Order (ants/bees/wasps)


```{r}
Hymnptr=commDat[commDat$Order=="Hymenoptera",] #This is read as "all columns of commDat where Order is Hymenoptera"
Hymnptr[,1:8]#let's just look at the first eight columns of this subsetted dataset
```

You can also do this with the "subset" function

```{r}
Hymnptr=subset(commDat,commDat$Order=="Hymenoptera")
Hymnptr[,1:8]
```

We can also just look at records for Freemont cottonwoods (fre)

```{r}
FremDat=commDat[,substr(colnames(commDat),1,3)=="fre"]
#let's breadk this down
colnames(commDat) #the column names of the dataset
substr(colnames(commDat),1,3) #selecting the first three letters of the column names
help(substr) #check out the parameters needed for substr
substr(colnames(commDat),1,3)=="fre" #this returns a TRUE/FALSE record of column names; TRUE if "fre", FALSE if not
head(FremDat)
```

what if we want to add the taxonomic info back into the Fremont dataset?

```{r}
FremDatTax=cbind(commDat[,1:6],FremDat)
head(FremDatTax)
```

We can still refer to specific taxonomic subsets of FremDat, even without the taxonomic data being in that dataset

```{r}
FremDatHymn=FremDat[commDat$Order=="Hymenoptera",] #this is read as "all columns of FremDat where Order is Hymenoptera"; the latter is indexed froma different object, commDat
FremDatHymn
```


##########
OK, now let's do something with our data
How many species occur in each sample (species richness)?
First we need to create a function

```{r}
SR=function(x){
return(length(x[x!=0])) #this is read as "return the number of observations in a sample that is not zero"
}
```

Now let's apply our function

```{r}
commSR=apply(commDat[,-c(1:6)],2,SR) #apply the SR function by columns (1=rows, 2=columns)
#let's break this down
head(commDat[,-c(1:6)]) #we only wanted to calculate species richness for the sample columns; thus, to apply our SR function, we said "apply this function to all columns except the first 6"
commSR #species richness per sample
```

Let's see how species richness varies across tree categories

```{r}
treeCat=c(rep(c("Fre","F1","BC","Narr"),each=10))
boxplot(commSR~treeCat,ylab="Species Richness")
```

Are there significant differences in SR among tree types?

```{r}
summary(aov(commSR~treeCat))#nope!
```

Let's save that species richness boxplot (Save it to your own folder, not the one below)

```{r}
tiff("U:/LOCKED/R_code/week03/Species_Richness_boxplot.tif",width=80,height=80,units="mm",res=150)
par(oma=c(0,0,0,0),mar=c(4,4,0,0))
boxplot(commSR~treeCat,xlab="Tree Type",ylab="Species Richness")
dev.off()
```


###########
Now let's do an analysis that is part of a package, vegan, that we have to download
1. Click on the "Packages" tab at right
2. Click the "Install" icon
3. Type in "vegan" (no quotes) and hit Enter

```{r}
library(vegan)
```

Let's look at a species accumulation curve for our Freemont trees to assess whether our sampling was sufficient

```{r}
help(specaccum) #let's look at the parameters and outputs for this function
FremAccum=specaccum(t(FremDat),method="random",perm=100)
#Let's break this down
t(FremDat) #specaccum wants the dataset with species as columns and communities as rows; t(FremDat) is the transpose of our Freemont sub-dataset
FremAccum
#This function randomly selects 1,2,...,10 samples 100 times, and calculates the mean and standard deviation of species richness
```

Let's plot richness as a function of samples

```{r}
plot(FremAccum$sites,FremAccum$richness,xlab="Samples",ylab="Richness") #looks like the species pool for Freemont was not fully sampled
#Let's connect the dots
lines(FremAccum$sites,FremAccum$richness)
```

Or we can be lazy and use in the built-in plot function for specaccum

```{r}
plot(FremAccum)
```

Let's see if we sufficiently sampled the "global" pool (species across all cottonwood types)

```{r}
AllAccum=specaccum(t(commDat[,-c(1:6)]),method="random",perm=100)
plot(AllAccum)
#Still not hitting an asymptote; should have sampled more...
```

Let's save this plot

```{r}
tiff("U:/LOCKED/R_code/week03/Species_Accum_Curve.tif",width=120,height=80,units="mm",res=150)
par(oma=c(0,0,0,0),mar=c(4,4,0,0))
plot(AllAccum)
dev.off()
```

That's it for this week.  Next week we'll start playing with some new tools for visualizing and analyzing community structure.